---
title: "R Example: Introduction to Dataframes"
output: html_notebook
---

This code was developed by: <br>
Shannon E. Albeke, Ph.D. <br>
University of Wyoming <br>
Wyoming Geographic Information Science Center <br>
Laramie, WY 82071 <br>
salbeke@uwyo.edu / 307-223-2129 <br>



```{r}
#Notice the colored text, when something is green, it is commented out and not executed. 
#Thus making comments to your code easy and very useful

#Let's learn a little bit about R objects and executing code
#simple arithmetic
2+2

#Store the value into a variable, be sure to look at your workspace. Notice we use the pipe "<-"
# instead of an equal sign "=". Use the pipe for variable declaration, equal for function arguments.
a<- 2+2

#create a list of numbers. This is also called a vector of type=numeric
b<- c(10, 20, 30, 40, 50)

#this is where R leverages a lot of power and has become very popular among researchers
( ab<- a*b )

#lets convert the numeric vector to something you may be more familiar with.
#data.frames are the R version of a spreadsheet or database table.
#Each column of a data.frame is considered a vector of a single data-type
dfv<- data.frame(col1=ab) #vertical table, 1 col, 5 rows

#You can look at the data stored in memory within the data.frame using this function
View(dfv) #or you can use the mouse and click on the object in your environment

#add 2 columns to the data frame using the column bind function
dfv<- cbind(dfv, col2=c(20, 30, 40, 50, 60), more=c(5:9))

#view the data.frame structure/data types. Notice that the column names are the
# same as we assigned within the arguments of the cbind function.
str(dfv)

#It's important to understand what happens when you convert/coerce from one data-type to another
#convert a column from numeric to factor and add results to new column
dfv[,4]<- as.factor(dfv[ , 2]) #Notice we are using the index for column, R will assign a column name
# The factor data-type is a categorical representation of unique values.
# Also notice that R will not name a column that begins with a number, thus V4, not just "4"
str(dfv)
#push col4 back to integer and into a new column
dfv$AsInt<- as.integer(dfv[ , 4])
dfv$AsInt
# Notice what happened to the original values when converting a factor to numeric!!
# To avoid the incorrect coersion, you must first convert to a character then numeric
dfv$AsInt<- as.integer(as.character(dfv[ , 4]))
dfv$AsInt #the values in column 4 are now back to their original values

#///////////////
#extract specific values from the 5 x 5 data frame
#row 1, col 3 USE INDEX
dfv[1,3]
#row 1, col 3 USE NAMES
names(dfv) #gives a list of column names
rownames(dfv) #gives a list of row names

dfv["1", "more"] #Notice names are characters, thus use of " " is necessary

```

```{r}
#once you have a data.frame with your data, you can perform all sorts of fun things, e.g.
#averages for each column of data

(colavg<- colMeans(dfv[,]) )

```
You will notice that the previous line threw an error. This is because column 4 is a factor. 
```{r}
#first remove col4
dfv<- dfv[,-c(4)]

#rerun the colMeans function
(colavg<- colMeans(dfv[,]) )

#NA means a NULL value, you have to be able to deal with them
dfv[1,3]<-NA

# In this example the 3rd column has an NA, thus R doesn't want to think for you by changing 
# the denominator and returns an NA
( colavg<- colMeans(dfv[,]) )

#By looking at the arguments of the colMeans function, you will see the na.rm argument
( colavg<- colMeans(dfv[,], na.rm = TRUE) )

#or ...sapply is a great function for iterating through an object, in this case several columns
( colavg2<- sapply(dfv, FUN=function(x){mean(x, na.rm=TRUE)}) ) #average
( colsd<- sapply(dfv, FUN=function(x){sd(x, na.rm=TRUE)}) ) #standard deviation
#or
summary(dfv)

#quick barplot...this is too simple, we will use ggplot2 for other plotting examples
barplot(colavg2)
#plot same date with confidence intervals
require(plotrix)
plotCI(colavg2, uiw=colsd, ylab="Random #", xlab="Column")

#As you can see, R can do a lot of things very quickly and easily that you have to take a little extra time to perform
#in software such as Excel, Access and ArcGIS. 
```

Now we will work with some real data and learn a couple of techniques for dealing with data that may not be given to you in an ideal format. One of the best lessons I can pass onto you is that when working with data, more often than not, if you have a dataset that is 1 row for each unique observation, data visualization and analyses will be much easier and efficient. Let's begin!

```{r}
#////////////////////////////////////////////////////
#Begin Analyses section
require(readxl)
require(ggplot2)

#set the working directory
setwd("C:/R_Intro")

#Read in the NDVI data from Ramesh ndvi. This requires the readxl package.
#We wrap the excel funtion and make sure the class is data.frame
dat<- data.frame(read_excel("ndvi_laramie_basin.xlsx"))
#Investigate our data
View(dat)
str(dat)
#plot the Normal data as lines
( p<- ggplot() + geom_line(data = dat, aes(x = Julian.Day, y = NDVI_Normal)) )

#plot second series, Dry
( p<- p + geom_line(data = dat, aes(x = Julian.Day, y = NDVI_Dry)) )

#The question is, are NDVI values significantly different between seasons?
# To test, reformat the data and run an anova to test for significant differences,
# this means adding a Normal/Dry category and moving the data from columns to rows
# By doing this we are creating a data.frame that is 1 observation per row, which
# is ideal in my mind. 

#Create a new data.frame the old fashioned way (we will demonstrate useful functions later)
#First get the Dry season data
ndvi<- data.frame(DOY = dat$Julian.Day, Season = "Dry", NDVI = dat$NDVI_Dry)
#append the Normal data rows
ndvi<- rbind(ndvi, data.frame(DOY = dat$Julian.Day, Season = "Normal", NDVI = dat$NDVI_Normal))

#make the same plot as before, but use the grouping variable so it is easier/faster!
( p<- ggplot() + geom_line(data = ndvi, aes(x = DOY, y = NDVI, group = Season, color = Season)) )

#Run the ANOVA
mod.aov<- aov(formula = NDVI ~ Season, data = ndvi)
summary(mod.aov)
#Run a Linear Regression
mod.lm<- lm(formula = NDVI ~ Season, data = ndvi)
summary(mod.lm)

#plot the NDVI as it relates to season, defaults to a box/whisker given season is a factor
plot(ndvi$Season, ndvi$NDVI)
#you will notice that there isn't a difference between seasons, but our eyes say different
# This is because Wyoming has winter, NDVI measures chlorophyll, thus limit data to 
# growing season.

#Now filter the data.frame to be summer time (DOY between 140 and 260)
# So we query which rows match values from a column
grow<- ndvi[ndvi$DOY >= 140 & ndvi$DOY <= 260,]

#Re-run the statistics
#Run the ANOVA
mod.aov<- aov(formula = NDVI ~ Season, data = grow)
summary(mod.aov)
#Run a Linear Regression
mod.lm<- lm(formula = NDVI ~ Season, data = grow)
summary(mod.lm)

#plot the NDVI as it relates to season
( p<- ggplot() + geom_line(data = grow, aes(x = DOY, y = NDVI, group = Season, color = Season)) )

#OK, great job! We have shown a difference. Now we can demonstrate a function
# based approach for modifying the data to be 1 row / observation
require(reshape2)

#use the melt function to create a vertical table
ndvi.melt<- melt(data = dat, id.vars = "Julian.Day", 
                 variable.name = "Season", value.name = "NDVI")

```

